---
title: "Demantia Prediction"
author: "Trevor Okinda"
date: "2024"
output:
  github_document: 
    toc: yes
    toc_depth: 4
    fig_width: 6
    fig_height: 4
    df_print: default
editor_options:
  chunk_output_type: console
---

# Student Details

|                                              |     |
|----------------------------------------------|-----|
| **Student ID Number**                        | 134780 |
| **Student Name**                             | Trevor Okinda |
| **BBIT 4.2 Group**                           | C |
| **Project Name**                             | A Rainfall prediction model |

# Setup Chunk

**Note:** the following KnitR options have been set as the global defaults: <BR> `knitr::opts_chunk$set(echo = TRUE, warning = FALSE, eval = TRUE, collapse = FALSE, tidy = TRUE)`.

More KnitR options are documented here <https://bookdown.org/yihui/rmarkdown-cookbook/chunk-options.html> and here <https://yihui.org/knitr/options/>.

```{r setup, include=FALSE}
library(formatR)
knitr::opts_chunk$set(
  warning = FALSE,
  collapse = FALSE
)
```

### Source: 

The dataset that was used can be downloaded here: *\<https://www.kaggle.com/datasets/shashwatwork/dementia-prediction-dataset\>*

### Reference:

*\<Tiwari, S. (2019). Dementia Prediction Dataset [Dataset]. Kaggle. https://www.kaggle.com/datasets/shashwatwork/dementia-prediction-dataset\>\
Refer to the APA 7th edition manual for rules on how to cite datasets: <https://apastyle.apa.org/style-grammar-guidelines/references/examples/data-set-references>*

# Exploratory Data Analysis
## Load dataset
```{r Load dataset}
# Load necessary libraries
library(tidyverse)

# Load dataset
dementia_data <- read.csv("demantia.csv", colClasses = c(
  Group = "factor",        # Demented/Nondemented
  Visit = "factor",       # Visit number
  MR_Delay = "numeric",    # MR Delay
  Gender = "factor",          # M/F (Gender)
  Hand = "factor",         # Hand (Handedness)
  EDUC = "numeric",        # Education
  SES = "factor",         # Socioeconomic Status
  MMSE = "numeric",        # Mini-Mental State Examination
  CDR = "numeric",         # Clinical Dementia Rating
  eTIV = "numeric",        # Estimated Total Intracranial Volume
  nWBV = "numeric",        # Normalized Whole Brain Volume
  ASF = "numeric"          # Atlas Scaling Factor
))


# Display the structure of the dataset
str(dementia_data)

# View the first few rows of the dataset
head(dementia_data)

# View the dataset in a separate viewer window
View(dementia_data)
```

## Measures of Frequency
```{r MOF}
# Measures of Frequency
# Frequency of categorical variables
group_freq <- table(dementia_data$Group)
visit_freq <- table(dementia_data$Visit)
gender_freq <- table(dementia_data$Gender)
hand_freq <- table(dementia_data$Hand)
ses_freq <- table(dementia_data$SES)
```

## Measures of Central Tendency
```{r MOCT}
# Measures of Central Tendency
# Mean, Median for numerical variables
mean_age <- mean(dementia_data$Age, na.rm = TRUE)
median_age <- median(dementia_data$Age, na.rm = TRUE)
mean_mmse <- mean(dementia_data$MMSE, na.rm = TRUE)
median_mmse <- median(dementia_data$MMSE, na.rm = TRUE)

# Mode function
get_mode <- function(v) {
  uniqv <- unique(v)
  uniqv[which.max(tabulate(match(v, uniqv)))]
}
mode_education <- get_mode(dementia_data$EDUC)
```

## Measures of Relationship
```{r MOR}
# Measures of Relationship
# Correlation matrix for numerical variables
cor_matrix <- cor(dementia_data %>% select(EDUC, MMSE, CDR, eTIV, nWBV, ASF), use = "complete.obs")

# Contingency table for categorical variables
group_gender_table <- table(dementia_data$Group, dementia_data$Gender)
```

## Print Results
```{r Print}
# Print results
print("Frequency of Group:")
print(group_freq)
print("Frequency of Visit:")
print(visit_freq)
print("Frequency of Gender:")
print(gender_freq)
print("Frequency of Handedness:")
print(hand_freq)
print("Frequency of SES:")
print(ses_freq)

print(paste("Mean Age:", mean_age))
print(paste("Median Age:", median_age))
print(paste("Mode Education:", mode_education))


print(paste("Mean MMSE:", mean_mmse))
print(paste("Median MMSE:", median_mmse))


print("Correlation Matrix:")
print(cor_matrix)

print("Contingency Table of Group and Gender:")
print(group_gender_table)
```

## ANOVA
```{r ANOVA}
# Load necessary libraries
library(tidyverse)
library(car)   # For Levene's Test
library(psych) # For skewness and kurtosis

# Perform ANOVA

# Perform ANOVA on MMSE by Group
anova_mmse <- aov(MMSE ~ Group, data = dementia_data)
summary(anova_mmse)

# Perform ANOVA on CDR by Group
anova_cdr <- aov(CDR ~ Group, data = dementia_data)
summary(anova_cdr)

# Check Assumptions for ANOVA
# 1. Homogeneity of Variances (Levene's Test)
leveneTest(MMSE ~ Group, data = dementia_data)
leveneTest(CDR ~ Group, data = dementia_data)

# 2. Normality of Residuals
shapiro.test(residuals(anova_mmse))
shapiro.test(residuals(anova_cdr))

# Post-hoc Analysis if ANOVA is significant
# Tukey's Honest Significant Difference (HSD) test
tukey_mmse <- TukeyHSD(anova_mmse)
tukey_cdr <- TukeyHSD(anova_cdr)

# Print results of post-hoc tests
print(tukey_mmse)
print(tukey_cdr)
```

## Plots
```{r Plots}
# Load necessary libraries
library(tidyverse)
library(GGally)  # For pairwise plots
library(reshape2) # For melting data for the heatmap

# Univariate Plots

# Histogram for EDUC (Education)
ggplot(dementia_data, aes(x = EDUC)) +
  geom_histogram(bins = 15, fill = "skyblue", color = "black") +
  labs(title = "Distribution of Education Level (EDUC)", x = "Education Level", y = "Frequency") +
  theme_minimal()

# Boxplot for MMSE by Group
ggplot(dementia_data, aes(x = Group, y = MMSE, fill = Group)) +
  geom_boxplot() +
  labs(title = "MMSE Scores by Group", x = "Group", y = "MMSE") +
  theme_minimal() +
  scale_fill_manual(values = c("Demented" = "orange", "Nondemented" = "green"))

# Bar plot for Gender Distribution
ggplot(dementia_data, aes(x = Gender)) +
  geom_bar(fill = "coral") +
  labs(title = "Gender Distribution", x = "Gender", y = "Count") +
  theme_minimal()

# Multivariate Plots

# Scatter plot of MMSE vs. eTIV, colored by Group
ggplot(dementia_data, aes(x = eTIV, y = MMSE, color = Group)) +
  geom_point(alpha = 0.6, size = 3) +
  labs(title = "Scatter Plot of eTIV vs. MMSE by Group", x = "eTIV (Total Intracranial Volume)", y = "MMSE") +
  theme_minimal() +
  scale_color_manual(values = c("Demented" = "red", "Nondemented" = "blue"))

# Pairwise plot for selected numerical variables
# Select only numeric columns for the pair plot
numeric_data <- dementia_data %>% select(MR_Delay, EDUC, MMSE, CDR, eTIV, nWBV, ASF)
ggpairs(numeric_data, aes(color = dementia_data$Group), title = "Pairwise Plot of Numeric Variables by Group")

# Correlation heatmap
# Calculate correlations only for numeric columns
cor_data <- cor(numeric_data, use = "complete.obs")
# Melt correlation data for ggplot
melted_cor_data <- melt(cor_data)
ggplot(melted_cor_data, aes(Var1, Var2, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0) +
  theme_minimal() +
  labs(title = "Correlation Heatmap", x = "Variable", y = "Variable") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Box plot for MMSE across SES levels
ggplot(dementia_data, aes(x = SES, y = MMSE, fill = SES)) +
  geom_boxplot() +
  labs(title = "MMSE Scores by Socioeconomic Status (SES)", x = "SES", y = "MMSE") +
  theme_minimal() +
  scale_fill_brewer(palette = "Pastel1")

```

# Preprocessing and Data Transformation
## Missing Values
```{r Missing Values}
# Load necessary library
library(tidyverse)

# Check for missing values in the dataset
# Summarize the number of missing values per column
missing_values_summary <- sapply(dementia_data, function(x) sum(is.na(x)))
print("Missing values per column:")
print(missing_values_summary)

# Display a summary of missing values across the dataset
total_missing_values <- sum(is.na(dementia_data))
cat("Total missing values in the dataset:", total_missing_values, "\n")
```

## Remove missing Values
```{r Remove Missing Values}
# Visualize missing values using a heatmap
library(ggplot2)
library(naniar) # For visualizing missing data

# Plot missing data pattern
gg_miss_var(dementia_data) +
  labs(title = "Missing Data Pattern", x = "Variables", y = "Missing Count")

# Show rows with missing values
missing_rows <- dementia_data[rowSums(is.na(dementia_data)) > 0, ]
print("Rows with missing values:")
print(missing_rows)

# Remove rows with missing values
dementia_data_clean <- na.omit(dementia_data)

# Confirm that there are no missing values in the cleaned dataset
sum(is.na(dementia_data_clean))  # Should return 0 if all missing values are removed

# Display the cleaned dataset
head(dementia_data_clean)

# Check the dimensions of the original vs cleaned dataset
cat("Original dataset dimensions:", dim(dementia_data), "\n")
cat("Cleaned dataset dimensions:", dim(dementia_data_clean), "\n")

```

# Training Model
## Data Splitting
```{r Data Splitting}
# Load necessary library
library(caret)

# Set seed for reproducibility
set.seed(123)

# Split the data: 70% for training, 30% for testing
train_index <- createDataPartition(dementia_data_clean$Group, p = 0.7, list = FALSE)
training_data <- dementia_data_clean[train_index, ]
testing_data <- dementia_data_clean[-train_index, ]

# Confirm the dimensions of the splits
cat("Training data dimensions:", dim(training_data), "\n")
cat("Testing data dimensions:", dim(testing_data), "\n")

# Display the first few rows of each split
head(training_data)
head(testing_data)
```

## Bootstrapping
```{r Bootstrapping}
# Load necessary library
library(boot)

# Define a function to calculate the statistic (e.g., mean MMSE)
mean_mmse <- function(data, indices) {
  sample_data <- data[indices, ]  # Resample data with the given indices
  return(mean(sample_data$MMSE, na.rm = TRUE))  # Return the mean of MMSE
}

# Set seed for reproducibility
set.seed(123)

# Perform bootstrapping
# Using 1000 resamples to estimate the mean of MMSE
bootstrap_results <- boot(data = dementia_data_clean, statistic = mean_mmse, R = 1000)

# Display results
print(bootstrap_results)

# Plot bootstrap results to visualize the distribution of the mean MMSE
plot(bootstrap_results)
```

## Cross-validation
```{r Cross-validation}
# Load necessary libraries
library(dplyr)
library(caret)
library(MLmetrics)
library(nnet)  # for multinom

# Drop 'Hand' column
dementia_data_clean <- dementia_data_clean %>% select(-Hand)

# Ensure 'Group' is a factor
dementia_data_clean$Group <- as.factor(dementia_data_clean$Group)

# Check levels (expecting 3)
levels(dementia_data_clean$Group)

# Set seed
set.seed(123)

# Define cross-validation settings
train_control <- trainControl(
  method = "cv",
  number = 10,
  classProbs = TRUE,
  summaryFunction = multiClassSummary,
  savePredictions = "final"
)

# Fit a multinomial logistic regression model
model_cv <- train(
  Group ~ ., 
  data = dementia_data_clean, 
  method = "multinom", 
  trControl = train_control,
  trace = FALSE  # suppress output
)

# Display results
print(model_cv)
print(model_cv$results)
```

## Training Different Models
```{r Different Models}
# Load necessary libraries
library(dplyr)
library(caret)
library(MLmetrics)

# Remove the 'Hand' column from the dataset
if ("Hand" %in% colnames(dementia_data_clean)) {
  dementia_data_clean <- dementia_data_clean %>% select(-Hand)
}


# Check the structure of the cleaned dataset
str(dementia_data_clean)

# Set seed for reproducibility
set.seed(123)

# Define cross-validation settings
train_control <- trainControl(
  method = "cv", 
  number = 10, 
  classProbs = TRUE,
  summaryFunction = multiClassSummary
)

# K-Nearest Neighbors Model
knn_model <- train(
  Group ~ ., 
  data = dementia_data_clean, 
  method = "knn", 
  trControl = train_control,
  tuneLength = 10  # Search for the optimal value of k
)

# Random Forest Model
rf_model <- train(
  Group ~ ., 
  data = dementia_data_clean, 
  method = "rf", 
  trControl = train_control
)

# Support Vector Machine Model
svm_model <- train(
  Group ~ ., 
  data = dementia_data_clean, 
  method = "svmRadial", 
  trControl = train_control
)

# Display results of each model
cat("K-Nearest Neighbors Model:\n")
print(knn_model$results)

cat("\nRandom Forest Model:\n")
print(rf_model$results)

cat("\nSupport Vector Machine Model:\n")
print(svm_model$results)
```

## Performance Comparison
```{r Performance Comparison}
# Combine the results of the models
model_list <- resamples(list(KNN = knn_model, RF = rf_model, SVM = svm_model))

# Summarize the results
summary(model_list)

# Plot the comparison of model performance
bwplot(model_list, metric = "Accuracy")  # You can change 'Accuracy' to other metrics like 'Kappa' if needed
```

## Saving Model
```{r Saving Model}
# Load the saved model
loaded_rf_model <- readRDS("./models/saved_rf_model.rds")

# Example new data for prediction (replace values with actual test data as needed)
new_data <- data.frame(
  Visit = factor(1),
  MR_Delay = 0,
  Gender = factor("M"),
  EDUC = 16,
  SES = factor(2),
  MMSE = 28,
  CDR = 0.5,
  eTIV = 1500,
  nWBV = 0.7,
  ASF = 1.05
)

# Use the loaded model to make predictions
predictions_loaded_model <- predict(loaded_rf_model, newdata = new_data)

# Print predictions
print(predictions_loaded_model)

```

